---
title: "Handling Census Data in R"
output: html_notebook
author: Michael Mann GWU mmann1123@gwu.edu
---

# Install
Before you begin this tutorial you must download R  [R Markdown](https://mirrors.nics.utk.edu/cran/). Then install [R Studio’s](https://rstudio.com/products/rstudio/download/) IDE (stands for integrated development environment), a powerful user interface for R. Get the Open Source Edition of RStudio Desktop. R Studio . Then open R Studio. 

If you have a pre-existing installation of R and/or RStudio, I highly recommend that you reinstall both and get as current as possible. It can be considerably harder to run old software than new.

# Libraries
R is an extensible system and many people share useful code they have developed as a package via CRAN and GitHub. This provides most of the functionality we are looking to use, for instance plotting, GIS operations, etc.  To install a package from CRAN, for example the dplyr package for data manipulation, here is one way to do it in the R console (there are others).

*Keep in mind you only need to install new packages (aka libraries) the first time you run R. Or if you want to update it to the most recent version.*
 
```{r   }
install.packages('dplyr') # data manipulation and cleaning

```

Now let's install the libraries (additional functions) we need to use for this tutorial. 

```{r  }
install.packages('sf')         # spatial data handling
remotes::install_github("walkerke/tidycensus")
  # census data download see https://walkerke.github.io/tidycensus/articles/basic-usage.html
install.packages("cli")        # needed for devtools install
install.packages('devtools')   # allows installation of libraries from github
devtools::install_github("tidyverse/ggplot2")    # plotting installing latest version from github so that it works with sf
          # if prompted please select "1" to install all additional packages
install.packages('ggthemes')   # plotting themes
```


# Getting Access to Census Data

Now that we have installed all the needed libraries (packages), we need to load them into R. *You will need to run this bit of code every time you open R* to get this to work. 

```{r  }
library(tidycensus)
library(sf)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(tidyverse)
options(tigris_use_cache = TRUE)
```


To get started working with tidycensus, users should load the package along with the tidyverse package, and set their Census API key. A key can be obtained from [Census API Key](http://api.census.gov/data/key_signup.html).  **It will provide you with a 40 digit text string. Please keep track of this number. Store it in a safe place.**
 
```{r include=T, eval=FALSE}
API_Key = 'enter_API_key'  # non working example - please paste your own in
census_api_key(API_Key, install = T, overwrite=TRUE)  
```
 
# Downloading DC Census Tract Data with Geometry

Now let's download both 5-year ACS data and decennial Census data for all census tracts in Washington DC with tract geometry for all available years.

## Part 1: 5-Year ACS Data (2009-2022)

```{r}
# Get available years for 5-year ACS data (2009-2024 as of latest release)
acs_available_years <- 2009:2023

# Initialize empty list to store ACS data for each year
dc_acs_data <- list()

# Download ACS data for each available year
for(year in acs_available_years) {
  tryCatch({
    # Download demographic and housing data for DC tracts with geometry
    dc_data <- get_acs(
      geography = "tract",
      state = "DC",           # Washington DC
      year = year,
      variables = c(
        total_pop = "B01003_001",      # Total population
        median_income = "B19013_001",   # Median household income
        white_alone = "B02001_002",     # White alone
        black_alone = "B02001_003",     # Black alone
        asian_alone = "B02001_005",     # Asian alone
        hispanic = "B03003_003",        # Hispanic or Latino
        # median_rent = "B25064_001",     # Median gross rent
        median_home_value = "B25077_001", # Median value (owner-occupied housing units)
        low_quart_rent = "B25057_001",   # Low quartile rent
        median_rent = "B25058_001",         # Median rent
        upper_quart_rent = "B25059_001",  # Upper quartile rent
        total_housing_units = "B25001_001", # Total housing units
        owner_occupied = "B25003_002",  # Owner-occupied housing units
        renter_occupied = "B25003_003"  # Renter-occupied housing units
      ),
      output = "wide",
      geometry = TRUE,        # Include tract geometry
      cache_table = TRUE,
      survey = "acs5"
    )
    
    # Add year column and data source
    dc_data$year <- year
    dc_data$data_source <- "ACS5"
    
    # Store in list
    dc_acs_data[[as.character(year)]] <- dc_data
    
    cat("Downloaded ACS data for year:", year, "\n")
    
  }, error = function(e) {
    cat("Error downloading ACS data for year", year, ":", e$message, "\n")
  })
}

# Combine all ACS years into one dataset
if(length(dc_acs_data) > 0) {
  dc_acs_all_years <- do.call(rbind, dc_acs_data)
  cat("Total ACS rows downloaded:", nrow(dc_acs_all_years), "\n")
  cat("ACS years available:", unique(dc_acs_all_years$year), "\n")
} else {
  cat("No ACS data was successfully downloaded\n")
}
```

## Part 2: Decennial Census Data ( 2000,  )

```{r}
# Available decennial census years with tract-level spatial data
# Note: 1990 not available via tidycensus API
decennial_years <- c(2000)

# Initialize empty list to store decennial data
dc_decennial_data <- list()

# Download decennial data for each available year
for(year in decennial_years) {
  tryCatch({
    
    # Variables vary by year due to questionnaire changes
    if(year == 2020) {
      # 2020 variables - use DHC (Demographic and Housing Characteristics File) variables
      variables <- c(
        total_pop = "P1_001N",         # Total population
        white_alone = "P1_005N",       # White alone
        black_alone = "P1_006N",       # Black or African American alone
        asian_alone = "P1_008N",       # Asian alone
        hispanic = "P2_002N",          # Hispanic or Latino
        total_housing_units = "H1_001N" # Total housing units
      )
      sumfile_param <- "pl"# "dhc"  # Demographic and Housing Characteristics File for 2020
    } else if(year == 2010) {
      # 2010 variables - use SF1 (Summary File 1)
      variables <- c(
        total_pop = "P001001",         # Total population
        white_alone = "P005003",       # White alone
        black_alone = "P005004",       # Black alone
        asian_alone = "P005006",       # Asian alone
        hispanic = "P004003",          # Hispanic or Latino
        total_housing_units = "H001001" # Total housing units
      )
      sumfile_param <- "sf1"  # Summary File 1 for 2010
    } else if(year == 2000) {
      # 2000 variables (SF1 for basic demographics, SF3 for detailed characteristics)
      variables <- c(
        total_pop = "P001001", # Total population
        white_alone = "P006002", # White alone
        black_alone = "P006003", # Black alone
        asian_alone = "P006005", # Asian alone
        hispanic = "P007010", # Hispanic or Latino
        total_housing_units = "H001001", # Total housing units
        median_income = "P053001", # Median household income (SF3)
        median_rent = "H063001", # Median gross rent (SF3)
        low_quart_rent = "H055001", # Low quartile rent
        upper_quart_rent = "H056001", # Upper quartile rent
        median_home_value = "H085001", # Median value owner-occupied (SF3)
        owner_occupied = "H004001" # Owner-occupied housing units
      )
      sumfile_param <- "sf3"  # Summary File 3 for 2000 to get income/housing data
    }
    
    dc_decennial <- get_decennial(
      geography = "tract",
      state = "DC",
      year = year,
      variables = variables,
      output = "wide",
      geometry = TRUE,
      sumfile = sumfile_param
    )
    
    # Add year column and data source
    dc_decennial$year <- year
    dc_decennial$data_source <- "Decennial"
    
    # Store in list
    dc_decennial_data[[as.character(year)]] <- dc_decennial
    
    cat("Downloaded decennial data for year:", year, "\n")
    
  }, error = function(e) {
    cat("Error downloading decennial data for year", year, ":", e$message, "\n")
  })
}

# Combine all decennial years
if(length(dc_decennial_data) > 0) {
  dc_decennial_all_years <- do.call(rbind, dc_decennial_data)
  cat("Total decennial rows downloaded:", nrow(dc_decennial_all_years), "\n")
  cat("Decennial years available:", unique(dc_decennial_all_years$year), "\n")
} else {
  cat("No decennial data was successfully downloaded\n")
}
```

## Part 3: Combine All Data Sources

```{r}
# Create a combined dataset with standardized column names
# Note: Due to different variable availability, we'll keep them separate but accessible

all_census_data <- list(
  acs = if(exists("dc_acs_all_years")) dc_acs_all_years else NULL,
  decennial = if(exists("dc_decennial_all_years")) dc_decennial_all_years else NULL
)

# Print summary
cat("=== DATA SUMMARY ===\n")
if(!is.null(all_census_data$acs)) {
  cat("ACS 5-year data available for years:", paste(sort(unique(all_census_data$acs$year)), collapse = ", "), "\n")
  cat("ACS observations:", nrow(all_census_data$acs), "\n")
}
if(!is.null(all_census_data$decennial)) {
  cat("Decennial data available for years:", paste(sort(unique(all_census_data$decennial$year)), collapse = ", "), "\n")
  cat("Decennial observations:", nrow(all_census_data$decennial), "\n")
}
```
 
 

# Write Out Shapefiles

Writing out data to shapefiles is also very easy. 

```{r}
# Export to GeoPackage with separate layers for each year
gpkg_filename <- "dc_census_tracts_complete.gpkg"

# Remove existing geopackage if it exists
if(file.exists(gpkg_filename)) {
  file.remove(gpkg_filename)
  cat("Removed existing geopackage\n")
}

# Export ACS data - each year as a separate layer
if(exists("dc_acs_data")) {
  for(year in names(dc_acs_data)) {
    layer_name <- paste0("acs_", year)
    # Set CRS to 4326 before writing
    dc_acs_data[[year]] <- st_transform(dc_acs_data[[year]], 4326)
    st_write(dc_acs_data[[year]], 
             dsn = gpkg_filename,  
             layer = layer_name,
             driver = "GPKG", 
             append = TRUE)
    cat("Added ACS layer:", layer_name, "\n")
  }
}

# Export decennial data - each year as a separate layer
if(exists("dc_decennial_data")) {
  for(year in names(dc_decennial_data)) {
    layer_name <- paste0("decennial_", year)
    st_write(dc_decennial_data[[year]], 
             dsn = gpkg_filename,  
             layer = layer_name,
             driver = "GPKG", 
             append = TRUE)
    cat("Added decennial layer:", layer_name, "\n")
  }
}
 

cat("Complete geopackage exported:", gpkg_filename, "\n")

# List all layers in the geopackage
if(file.exists(gpkg_filename)) {
  layers <- st_layers(gpkg_filename)
  cat("Geopackage contains", length(layers$name), "layers:\n")
  for(i in 1:length(layers$name)) {
    cat(" -", layers$name[i], "(", layers$features[i], "features )\n")
  }
}
 
```


